Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-11-04 paper/pool_model.tex
2nd part of shared session
Co-authored-by: Polina Gaindrik <polinagaindrik@users.noreply.github.com>

Another mechanism is the motility of bacteria, e.g.,, the flagellates \textit{E. coli} or

@@

These effects can only be treated with a spatial model that is able to describe the observed

@@

To this end, we consider the system consisting from two bacterial species that can be described by two internal states: lag and growth states.

@@

$\alpha$ and different lengths of lag phase $\lambda$.
Moreover, the species $A$ is able to release a certain antimicrobial component $I$ that shows an
inhibitory effect on species 2 (amensalism).

@@

The \acp{ode} the bacterial population can be written as:

@@

    \dot{G_B} &= \lambda_B R L_B + \frac{\alpha_B}{1 + \mu_I I} R G_B

@@

The microenvironment of our system contains the resource and the inhibitor molecules:

@@

@@

% ------------------------------------------------------------------------------------------------
\subsection{Constructing an agent-based pool model}
The following sections will demonstrate how a given \ac{ode} system can be translated into a
custom-built \ac{abm} which has only a few more overall parameters.
This procedure is a blueprint which can be followed to construct \acp{abm} for similar
\ac{ode}-based systems.\\
To construct our \ac{abm}, we use \texttt{cellular\_raza}~\citep{Pleyer2025}.
It allows us to define our bacterial agents from a clean slate, without having to rely on and inherit
existing work.
With this flexibility, researchers are able to construct minimal models using only as many parameters which are necessary.
This ability, to precisely define and control the exact behavior of our cell-agents, enables the following parts.

@@

The first is uptake of an extracellular resource $[R_e]$ into the cell $[R_i]$ with rate $u$ in \unit{\per\minute}.
Nutrients are only taken up at the position of the cell $x_c$
    \dot{[R_i]} &= u [R_e](x_c)\\
    \dot{[R_e]}(x) &= -u \frac{V_c}{V_D} [R_e] \delta(x-x_c)
where $\delta(x-x_\text{cell})$ is the Dirac delta-distribution.

@@

In the second equation, we need to account for the ratio of cellular $V_c$ and domain volume $V_D$ to calculate the correct change.
Further, we want to modify the cellular volume by growing it increasingly with rate $a$.

@@

The total amount of resource molecules present inside the cell is given by the current concentration multiplied by the cell-volume $V_c [R_i]$.

@@

    \dot{V_c} &= a\sigma [R_i] V_c\\
    \dot{[R_i]} &= - a[R_i]
Furthermore, we let cells produce an inhibitor with production rate $b$ which inhibits the cell's ability to uptake and thus use the resource by a factor of $1+\nu_I [I]$
    \dot{[I]} &= b\frac{V_c}{V_D}\delta(x-x_c)\\
    u &\longrightarrow \frac{u}{1+\nu_I [I]}.
To distinguish between the two species, it suffices to set parameters for every individual agent accordingly.
This means, we can implement the most general version of the equations and by choosing appropriate parameters, this defines the species of the respective cell-agent.
If we combine all equations and consider multiple cells, we obtain the system
    \dot{V_c} &= a\sigma [R_i] V_c\\
    \dot{[R_i]} &= \frac{u}{1+\nu_I [I]} [R_e] - a[R_i]\\
    \dot{[R_e]} &= -\frac{u}{1+\nu_I [I]} \frac{V_c}{V_D} [R_e] \delta(x-x_c)\\
    \dot{[I]} &= b\frac{V_c}{V_D}\delta(x-x_c).
    \label{eq:spatial-full-system-non-simplified}

@@

This means, we can effectively set $\dot{[R_i]}=0$.
Therefore, we obtain the relation $\frac{u}{1+\nu_I [I]}[R_e] = a[R_i]$ which we can use to eliminate the variable monitoring internal resource concentrations $[R_i]$.
Additionally, we require our external components to be freely diffusible with constants $D_R,D_I$.
Plugging this into equation~\eqref{eq:spatial-full-system-non-simplified} yields.
    \dot{V_c} &= \frac{u}{1+\nu_I [I]} \sigma [R]V_c\\
    \dot{[R]} &= \sum\limits_c -\frac{u}{1+\nu_I [I]} \frac{V_c}{V_D} [R] \delta(x-x_c) + D_R\Delta [R]\\
    \dot{[I]} &= \sum\limits_c b\frac{V_c}{V_D}\delta(x-x_c) + D_I\Delta [I].

@@

Since, we will not explicitly model intracellular concentrations $[R_i]$, we will drop the index entirely, only considering the extracellular resource.

@@

We represent cells as spherical objects in 2D space.
They can be represented by their position $x_c$ and radius $r_c$ which is of course related to their (2D) volume $V_c = \pi r_c^2$.

@@

Notice however, that how the radius is updated does not correspond to the cellular reactions defined previously but is a purely mechanical consideration.

@@

The underlying equation of motion is that of a standard damped particle
    m \ddot{\vec{x}} = -\eta\dot{\vec{x}} + \vec{F}

@@

These mechanisms will result in the overall expansion of the colony as individual cells push each further other apart.

@@

Cell growth is already being described the system of coupled \acp{ode} given in equations~\eqref{eq:spatial-full-system-simplified}.

@@

The \ac{abm} model should be considered to be the more fundamental model compared to the \ac{ode} model in the sense that it incorporates more biological aspects and maps them closer to the actual behaviour of the bacteria.

@@

>>>>

Another mechanism is the motility of bacteria, e.g., the flagellates of \textit{E. coli} or

@@

These effects can only be treated within a spatial framework that is able to describe the observed

@@

To this end, we consider the system consisting of two bacterial species that can be described by the two internal states: lag phase and growth.

@@

$\alpha_{A/B}$ and different lag phase transition rates $\lambda_{A/B}$.
Moreover, species $A$ is able to release a certain antimicrobial component $I$ that shows an
inhibitory effect on species $B$ (amensalism).

@@

The \acp{ode} system of the bacterial population can be written as:

@@

        \dot{G_B} &= \lambda_B R L_B + \frac{\alpha_B}{1 + \mu_I I} R G_B.

@@

Moreover, the microenvironment of our system contains the resource and the inhibitor molecules

@@

        \dot{I} &= \kappa G_A.

@@

% ------------------------------------------------------------------------------------------------
\subsection{Constructing an agent-based pool model}
The following sections will demonstrate how a given \ac{ode} system can be translated into a
custom-built \ac{abm} which has only a few more overall parameters.
This procedure is a blueprint which can be followed to construct \acp{abm} for similar
\ac{ode}-based systems.\\
To construct our \ac{abm}, we use the modeling framework \texttt{cellular\_raza}~\citep{Pleyer2025}.
 allows us to define our bacterial agents from a clean slate.
With this flexibility, researchers are able to construct minimal models using only as many parameters which are necessary.
This ability, to precisely define and control the exact behavior of our cell-agents, enables the following parts.

@@

We denote cells of species $A,B$ with $c_{A/B}$ respectively and indicate shared behavior by simply using $c$.
Cells $c_A$ of species $A$ take up extracellular resources $[R_e]$ into the cell $[R_{i,c_A}]$ with rate $u$ in \unit{\per\minute}.
The uptake only occurs at the position of the cell $x_{c_A}$.
Furthermore, we let cells of species $A$ produce a freely diffusable inhibitor $I$ with production rate $b$ and diffusion constant $D_I$ which inhibits cells of species $B$ in their uptake of nutrients.
    \dot{[I]} = b\frac{V_{c_A}}{V_D}\delta(x-x_{c_A}) + D_I \Delta [I]
We model this inhibiting effect by modifying the uptake term of species $B$ with a factor of $\frac{1}{1+\nu_I [I]}$.
% To distinguish between the two species, it suffices to set parameters for every individual agent accordingly.
% This means, we can implement the most general version of the equations and by choosing appropriate parameters, this defines the species of the respective cell-agent.
If we combine these effects and consider multiple cells, we obtain the system
    \dot{[R_{i,c_A}]} &=& u &[R_e] - a[R_{i,c_A}]\\
    \label{eq:spatial-intracellular-a}
    \dot{[R_{i,c_B}]} &=& \frac{u}{1+\nu_I [I]} &[R_e] - a[R_{i,c_B}]\\
    \label{eq:spatial-intracellular-b}
    % TODO fix this
    % \begin{split}
        \dot{[R_e]} &=& u &\sum\limits_{c_A}\frac{V_{c_A}}{V_D} [R_e] \delta(x-x_{c_A})\\
            &&-\frac{u}{1+\nu_I [I]} &\sum\limits_{c_B}\frac{V_{c_B}}{V_D} [R_e] \delta(x-x_{c_B})
    % \end{split}
    \label{eq:spatial-extracellular-resources-pde}
where equations~\eqref{eq:spatial-intracellular-a} and \eqref{eq:spatial-intracellular-b} describe the change of intracellular resources for every individual cell of of species $A,B$ and the equation~\eqref{eq:spatial-extracellular-resources-pde} contains extracellular resources coupled to each individual agent.
They contain the uptake of resources and term used grow the cell, converting resources into volume which will be explained below.
We use the Dirac delta-distribution $\delta(x-x_c)$ to enforce exchange of nutrients at the positions of the cells $x_{c}$.

@@

In equation~\eqref{eq:spatial-extracellular-resources-pde} we need to account for the ratio of cellular $V_c$ and domain volume $V_D$ to calculate the correct exchange.
In addition, we want to modify the cellular volume by growing it increasingly with rate $a$.

@@

The total amount of resource molecules present inside the cell is given by the current concentration multiplied by the cell-volume $V_c [R_{i,c}]$.

@@

    \dot{V_c} = a\sigma [R_{i,c}] V_c

@@

This means, we can effectively set $\dot{[R_{i,c}]}=0$.
Therefore, we obtain the relations $\frac{u}{1+\nu_I [I]}[R_e] = a[R_{i,cB}]$ and $u[R_e]=a[R_{i,c_A}]$ which we can use to eliminate the variable monitoring internal resource concentrations $[R_{i,c}]$.
Additionally, we also require the nutrients to be freely diffusible with constants $D_R$.
With these considerations, we can update equations~\eqref{eq:spatial-intracellular-a} to \eqref{eq:spatial-extracellular-resources-pde} and obtain:
        % TODO format these equations
        \dot{V_{c_B}} &= u \sigma [R]V_c\\
        \dot{V_{c_A}} &= \frac{u}{1+\nu_I [I]} \sigma [R]V_c\\
        % \dot{[R_e]} &= \sum\limits_c -\frac{u}{1+\nu_I [I]} \frac{V_c}{V_D} [R] \delta(x-x_c) + D_R\Delta [R]\\
        \dot{[R]} &= u \sum\limits_{c_A}\frac{V_{c_A}}{V_D} [R_e] \delta(x-x_{c_A})\\
            &-\frac{u}{1+\nu_I [I]} \sum\limits_{c_B}\frac{V_{c_B}}{V_D} [R_e] \delta(x-x_{c_B})\\
            &+D_R \Delta [R_e]\\
        \dot{[I]} &= \sum\limits_c b\frac{V_c}{V_D}\delta(x-x_c) + D_I\Delta [I]

@@

Since, we will not explicitly model intracellular concentrations $[R_{i,c}]$, we will drop the index entirely, only considering the extracellular resource.

@@

We model cells as spherical objects in 2D space.
They can be represented by their position $x_c$ and radius $r_c$ which is related to their (2D) volume $V_c = \pi r_c^2$.

@@

% Notice however, that how the radius is updated does not correspond to the cellular reactions defined previously but is a purely mechanical consideration.

@@

The dynamics are described by newtondian motions for particle damping.
    m \ddot{\vec{x}} = -\eta\dot{\vec{x}} + \vec{F}.

@@

These mechanisms will result in the overall expansion of the colony as individual cells push each other further apart.

@@

Cell growth is already being described by the system of coupled \acp{ode} given in equations~\eqref{eq:spatial-full-system-simplified}.

@@

The \ac{abm} model should be considered to be more fundamental compared to the \ac{ode} description in the sense that it incorporates more biological aspects and maps them closer to the actual behaviour of the bacteria.

@@

    \label{tab:mappings-ode-abm}
